%begin-include

\unsection{Exercises}

\begin{exercises}
\item The exclusive-or connective acts on two propositional variables producing a propositional form $p\otimes q$ that is equivalent to $(p\land \lnot q) \lor (\lnot p \land q)$. Write down the truth table of this connective.

\item In a parallel universe, all mathematicians have assembled and formed an organization known as the Chalk Party. One member of the Chalk Party is being judged for allegedly having stolen chalk from the University of Oviedo.

When the prosecutor begins his exposition, he says: ``This man is clearly guilty. He is a member of the chalk party, and all chalk thieves belong to this organization''. The judge is quick to dismiss his argument as a fallacy. Why? Justify your answer using propositional logic. 

\item Prove that the empty set is a subset of any set $A$.

\emph{Hint:} Show that, if $x\in \emptyset$, then $x\in A$.

\item Generally, any proof relying on an algorithm can be easily transformed into a proof by induction. Rewrite the proof of \ref{ads1} as a proof by induction.

\item Find an equivalent expression for each of the following predicates in which any negation connectives are acting directly on $P$, $Q$ or $R$.
\begin{lettered}
\item $\lnot (\forall x)(\exists y)(\forall z)( (P(x) \land Q(y))\lor (R(z))$.
\item $\lnot (\exists x)(\forall y) (\forall z) ( P(x) \limplies (Q(y) \land R(z)))$.
\end{lettered}

\item Use induction to prove that, given a natural number $n$ and some propositional variables $p, q_1,\ldots,q_n$, the following propositional forms are tautologies.
\begin{lettered}
\item $p \land (q_1 \lor \cdots \lor q_n) \liff (p\lor q_1) \land \cdots \land (p \lor q_n)$.
\item $p \lor (q_1 \land \cdots \land q_n) \liff (p\land q_1) \lor \cdots \lor (p \land q_n)$.
\item $\lnot(q_1\lor \cdots \lor q_n) \liff \lnot q_1 \land \cdots \land \lnot q_n$. 
\item $\lnot(q_1\land \cdots \land q_n) \liff \lnot q_1 \lor \cdots \lor \lnot q_n$. 
\end{lettered}


\item Use induction and double inclusion to prove that, given a natural number $n$ and some sets $A, B_1,\ldots,B_n$, we have
\begin{lettered}
\item $A \cap (B_1 \cup \cdots \cup B_n) = (A\cup B_1) \cap \cdots \cap (A \cup B_n)$,
\item $A \cup (B_1 \cap \cdots \cap B_n) = (A\cap B_1) \cup \cdots \cup (A \cap B_n)$.
\end{lettered}
Let $A_1,\ldots,A_n$ be subsets of a set $X$. Show that
\begin{lettered}[resume]
\item $X \setminus (A_1 \cup \cdots \cup A_n) = (X \setminus A_1) \cap \cdots \cap (X \setminus A_n)$,
\item $X \setminus (A_1 \cap \cdots \cap A_n) = (X \setminus A_1) \cup \cdots \cup (X \setminus A_n)$.
\end{lettered}
\end{exercises}
